---
title: "Multimodally profiling memory T cells from a tuberculosis cohort identifies cell state associations with demographics, environment and disease (GSE158769)"
subtitle: '[Nathan, et al. Nat Immunol  (2021)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE158769)'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
    toc: true
    smart: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  cache = TRUE,
  cache.lazy = FALSE)
```

Perform dreamlet analysis of single cell data assaying T cell populations in donors infected with tuberculosis.

# Loading libraries
```{r load.packages, cache=FALSE}
library(SingleCellExperiment)
library(zellkonverter)
library(ggplot2)
library(scattermore)
library(crumblr)
library(aplot)
library(tidyverse)
library(dreamlet)
library(kableExtra)
library(ggcorrplot)
library(RColorBrewer)
library(DelayedArray)

setAutoBlockSize(1e9)
```

# Load data
```{r load.data}
# read H5AD file
path = "/sc/arion/projects/CommonMind/hoffman/scRNAseq_data/Nathan_NatImm_2021/"
file = paste0(path, "/Nathan_NatImm_2021.h5ad")
sce = readH5AD(file, use_hdf5=TRUE)   

# get only single-cell RNA-seq
# rest is CITE-seq proteins
only_rna = grep("prot", rownames(sce), invert=TRUE)

# create variable that distinguishes cells from the same donor
# but different batches
sce$donor_batch = with(colData(sce), paste(donor, batch))
sce$batch = factor(sce$batch)
sce$season = factor(sce$season)
sce$TB_status = factor(sce$TB_status, c("CONTROL", "CASE"))

reducedDim(sce, "UMAP") = with(colData(sce), data.frame(UMAP_1, UMAP_2))

# sort cell clusters by name
sce = sce[only_rna,sce$cluster_name != "NA"]
colData(sce) = droplevels(colData(sce))
tab = colData(sce) %>%
  as_tibble %>%
  group_by(cluster_ids, cluster_name) %>%
  filter(row_number()==1) %>%
  summarise(cluster_ids, cluster_name) %>%
  arrange(as.numeric(gsub("C-", '', cluster_ids)))

sce$cluster_name = factor(sce$cluster_name, unique(tab$cluster_name))
sce$cluster_ids = factor(sce$cluster_ids, unique(tab$cluster_ids))
```


## Plot UMAP
```{r UMAP}
colData(sce) %>%  
  data.frame %>%
  ggplot(aes(UMAP_1, UMAP_2, color=cluster_name)) +
    geom_scattermore() +
    theme_classic() +
    theme(aspect.ratio=1)
```

## Create pseudobulk
```{r pb}
pb <- aggregateToPseudoBulk(sce,
    assay = "counts",     
    cluster_id = "cluster_name",  
    sample_id = "donor_batch")
```

# Plots of cell fractions
```{r fractions, fig.height=11, fig.width=11}
fracs = cellCounts(pb) / rowSums(cellCounts(pb))

df = data.frame(fracs, TB_status = pb$TB_status, check.names=FALSE)
df = reshape2::melt(df, id.vars="TB_status")

ggplot(df, aes(TB_status, value, fill=variable)) + 
  geom_violin() +
  geom_boxplot(fill="grey50", width=.1) + 
  facet_wrap(~ variable) +
  theme_classic() +
  theme(aspect.ratio=1, legend.position="none") +
  ylab("Fraction")
```

# Evaluate crumblr and build tree
```{r test.crumblr}
# run crumblr on cell counts
cobj = crumblr(cellCounts(pb))

# build tree from pseudobulk expression
hcl = buildClusterTreeFromPB( pb )
```


## Variance partitioning analysis
Scale `age` to avoid numerical issues.
```{r vp}
form = ~ (1|TB_status) + (1|donor) + (1|batch) + (1|season) + (1|sex) + scale(age) + I(scale(age)^2) + scale(prop_NAT)

vp = fitExtractVarPartModel(cobj, form, colData(pb))

vp$age = vp[,'scale(age)'] + vp[,'I(scale(age)^2)']
vp[,'scale(age)']  = c()
vp[,'I(scale(age)^2)'] = c()

cols = c(brewer.pal(ncol(vp)-1, "Set1"), "grey85")

plotVarPart(sortCols(vp), col=cols)
```

```{r plotPercentBars1}
plotPercentBars(sortCols(vp), col=cols) 
```

## Removing repeated measures
Keep only a single sample per donor
```{r vp.nodup}
form = ~ (1|TB_status) + (1|batch) + (1|season) + (1|sex) + scale(age) + I(scale(age)^2) + scale(prop_NAT)

keep = !duplicated(colData(pb)$donor)
vp = fitExtractVarPartModel(cobj[,keep], form, colData(pb)[keep,])

vp$age = vp[,'scale(age)'] + vp[,'I(scale(age)^2)']
vp[,'scale(age)']  = c()
vp[,'I(scale(age)^2)'] = c()

cols = c(brewer.pal(ncol(vp)-1, "Set1"), "grey85")

plotVarPart(sortCols(vp), col=cols)
```


```{r plotPercentBars2}
fig.vp = plotPercentBars(sortCols(vp), col=cols) 
fig.vp
```


## Differential testing
```{r diff.test}
# Run precision-weighted linear mixed model using dream()
# and compute empirical Bayes moderated t-statistics
form = ~ TB_status + (1|donor) + (1|batch) + (1|season) + (1|sex) + scale(age) + I(scale(age)^2) + scale(prop_NAT)
fit = dream(cobj, form, colData(pb))
fit = eBayes(fit) 

# show results
topTable(fit, coef='TB_statusCASE', number=Inf) %>%   
  select(logFC, AveExpr, t, P.Value, adj.P.Val) %>% 
  kbl() %>%  
  kable_classic(full_width = FALSE)
```

## Multivariate test along hierarchy
```{r treeTest, fig.height=6, fig.width=12}
# Perform multivariate testing 
res1 = treeTest( fit, cobj, hcl, coef="TB_statusCASE")

# plot tree with results
fig.tree = plotTreeTest(res1) + xlim(0, 15) + theme(legend.position="none")

# plot of logFC
tab = topTable(fit, "TB_statusCASE", number=Inf, sort.by="none")
tab$celltype = factor(rownames(tab), rev(get_taxa_name(fig.tree)))
tab$se = with(tab, logFC/ t)

fig.logFC = ggplot(tab, aes(celltype, logFC)) +
  geom_hline(yintercept=0, linetype="dashed", color="grey50") + 
  geom_errorbar(aes(ymin=logFC - 1.96*se, ymax=logFC + 1.96*se), width=0) +
  geom_point(color="dodgerblue") +
  coord_flip() +
  theme_classic() +
  xlab('') +
  theme(aspect.ratio=3, axis.text.y = element_blank())
 
# combine plots
fig.logFC %>% insert_left(fig.tree)  %>% insert_right(fig.vp)
```


## Correlation between coefficient estimates
```{r ggcorrplot, fig.height=8, fig.width=8}
# get covariance matrix
V = vcov(fit, cobj, coef="TB_statusCASE")

# convert to correlation matrix
V = cov2cor(V)

# shorten names
rownames(V) = gsub(":TB_statusCASE", '', rownames(V))
colnames(V) = gsub(":TB_statusCASE", '', colnames(V))

# sort to match tree
i = match(rev(get_taxa_name(fig.tree)), rownames(V))

ggcorrplot(V[i,i], outline.color=NA, colors=c("blue3", "white", "red3"))
```






# Session info
<details>
```{r sessionInfo}
sessionInfo()
```
</details>














