---
title: "Test differences in cell type composition"
subtitle: ''
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
---



<!---
cd /Users/gabrielhoffman/workspace/repos/crumblr_analysis/simulations
R

rmarkdown::render('simulations.Rmd')

# module purge
# ml anaconda3/2024.02
# conda create -n "sccoda" python 

# new screen
conda activate sccoda
# pip install sccoda

ml pandoc
ENV=/sc/arion/work/hoffmg01/condaEnv/sccoda/
conda activate $ENV
export LD_LIBRARY_PATH=$ENV/lib/:$LD_LIBRARY_PATH

cd /sc/arion/projects/CommonMind/hoffman/crumblr_analysis/simulations
R

system("git pull"); rmarkdown::render("simulations.Rmd");


# https://hoffmg01.hpc.mssm.edu/crumblr_analysis/simulations/simulations.html

--->

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(	
						tidy=FALSE, 
						cache=TRUE,
						echo=FALSE,
          	dev=c("png", "pdf"),
          	# dev.args = list(png = list(type = "Xlib")),
          	package.startup.message = FALSE,
          	message=FALSE, 
          	error=FALSE, 
          	warning=FALSE)

knitr::opts_chunk$set()
```	


```{R load.packages, cache=FALSE}
library(crumblr)
library(variancePartition)
library(PRROC)
library(MASS)
library(compositions)
library(parallel)
library(Rfast)
library(cowplot)
library(HMP)
library(gtools)
library(RColorBrewer)
library(tidyverse)

library(reticulate)
python_path = "/sc/arion/work/hoffmg01/condaEnv/sccoda/bin/python"
use_python(python_path)
scoda = import("sccoda")
scoda.utils = import("sccoda.util.cell_composition_data")
scoda.analysis = import("sccoda.util.comp_ana")

source("scCoda_wrapper.R")
source("cellTypeCompositionTest.R")
```



```{r sim.functions}
rmultinomdir = function(n, size, alpha){

	p = rdirichlet(n, alpha)
	res = lapply( seq(1, n), function(i){
		rmultinom(1, size, prob=p[i,])
		})
	res = do.call(cbind, res)

	t(res)
} 

run_simulation = function( n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, formula, methods, tau, use.sccoda=FALSE, mc.cores=20){

	# each batch is observed at last twice
	info = data.frame(trait = rnorm(n_samples), Batch = factor(rep(sample.int(n_batches, n_samples/2, replace=TRUE), 2)))

	rownames(info) = paste0("sample_", 1:nrow(info))

	df_res = mclapply( 1:n_sims, function(idx){

		RhpcBLASctl::omp_set_num_threads(1)

		if( idx %% 1 == 0) message("\r", idx, " / ", n_sims, '    ', appendLF=FALSE)
		set.seed(idx)

		value = ifelse( idx <= n_sims*0.9, 0, beta.true )
		beta = data.frame(mu = rep(0, n_clusters), b = c(0, value, rep(0, n_clusters-2)))
		a = n_clusters 
		b = nlevels(info$Batch)-1
		beta_batch = matrix(rnorm(a*b, sd=sqrt(var.batch)), a, b)
		beta = cbind(beta, beta_batch)

		eta = model.matrix(~., info) %*% t(beta)

		probs = exp(eta)

		totalCounts = rnegbin(nrow(probs), mu=mean_counts, theta = 3)

		if( tau == 1){
			# large a0 corresponds to multinomial
			alpha = probs * 1e9
		}else{
			# convert tau overdispersion value to alpha from p
			rhoSq = (tau - 1) / (mean(totalCounts)-1)
			a0 = (1-rhoSq) / rhoSq
			alpha = probs*a0
		}
 
		# Multinomial-Dirichlet
		counts = lapply( 1:length(totalCounts), function(i){
			Dirichlet.multinomial(totalCounts[i], alpha[i,]) 
		})
		counts = do.call(rbind, counts)
		rownames(counts) = paste0("sample_", 1:nrow(counts))
		colnames(counts) = paste0("CT_", 1:ncol(counts))

		cellTypeFreq = rowMeans(apply(counts, 1, function(x) x/sum(x)))
		df_cellTypeFreq = data.frame(cellTypeFreq)

		# CLR
		Y = t(as.data.frame(compositions::clr(counts)))
		fit_clr = suppressMessages(dream( Y, formula, info, BPPARAM=SerialParam(), useWeights=FALSE))
		fit_clr = eBayes(fit_clr)
		tab_clr = topTable(fit_clr, coef="trait", sort.by="none")

		# crumblr
		cobj = crumblr(counts, max.ratio=5)
		fit = suppressMessages(dream(cobj, formula, info, BPPARAM=SerialParam()))
		fit = eBayes(fit)
		tab.crumblr = topTable(fit, coef="trait", sort.by="none")

		# # crumblr clr_2class
		# cobj2cl = crumblr(counts, method="clr_2class", max.ratio=5)
		# fit2cl = suppressMessages(dream(cobj2cl, formula, X, BPPARAM=SerialParam()))
		# fit2cl = eBayes(fit2cl)
		# tab.crumblr2cl = topTable(fit2cl, coef="trait", sort.by="none")
			# "crumblr2cl" = tab.crumblr2cl$P.Value,

		df_p = data.frame(assay = rownames(tab_clr),
			"CLR" = tab_clr$P.Value,
			check.names=FALSE)
		i = match(df_p$assay, rownames(tab.crumblr))
		df_p$crumblr = tab.crumblr$P.Value[i]

		if( use.sccoda){
			res = sccoda_wrapper(counts, info, formula, python_path = "/sc/arion/work/hoffmg01/condaEnv/sccoda/bin/python")

			res$scCoda = 1 - res[,'Inclusion probability']  

			df_p = merge(df_p, res[res$variable == "trait",c('scCoda','response')], by.x="assay", by.y="response") 
			rownames(df_p) = df_p$assay
			df_p = df_p[rownames(tab_clr),]
		}

		df_methods = lapply( methods, function(method){
			
			if( length(findbars(formula)) > 0 & method == "betabinomial"){
				# betabinomial can't accept random effects
				
				res = data.frame(Method = method, 
								assay = "CT_1",   
								Estimate = NA,  
								se = NA,    
								zstat  = NA,
								pValue = NA, 
								p.adj= NA)
			}else{

				# keep only samples with at elast 10 reads
				i = rowSums(counts) > 10
				res = testComposition( counts[i,], formula, info[i,], coef="trait", eval="test", method = method)
				res = data.frame(Method = method, res, check.names=FALSE)
			}
			res
		})
		df_methods = do.call(rbind, df_methods)

		df = with(df_methods, data.frame(Method, assay, p.value = pValue, check.names=FALSE))
		df = rbind(df, reshape2::melt(df_p, value.name="p.value", variable.name="Method", id.vars="assay"))

		df$i = idx
		df$value = value

		df = merge(df, df_cellTypeFreq, by.x="assay", by.y="row.names")

		save(df, counts, info, file = paste0("/sc/arion/scratch/hoffmg01/sims_", idx, ".RDATA"))

		df
	}, mc.cores=mc.cores)
	df_res = do.call(rbind, df_res)

	if( !is.numeric(df_res$p.value) ) save.image()

	# small offset to fix error if p-value is exactly zero
	df_res$p.value <- df_res$p.value + 1e-300

	# Compute AUPR
	df_aupr = lapply( unique(df_res$Method), function(method){

		res1 = lapply( unique(df_res$assay), function(CT){

			df = df_res[(df_res$Method == method)&(df_res$assay == CT),]

			df = df[!is.na(df$p.value),]

			if( nrow(df) == 0){
				res = data.frame(method,  
					assay = CT, 
					 AUPR = 0, AUPR.rnd=NA,
					ROC = 0, ROC.rnd = NA, FPR = 0)
			}else{
				isZero = df$value == 0

				score0 = -log10(df$p.value[!isZero])
				score1 = -log10(df$p.value[isZero])

				if( method == "scCoda") score0 = rnorm(length(score0), 0, 1e-6)
				pr = pr.curve(score0, score1, curve=TRUE, rand.compute=TRUE)
				roc = roc.curve(score0, score1, curve=TRUE, rand.compute=TRUE)

				res = data.frame(method, 
					assay = CT,
					AUPR = pr$auc.integral, 
					AUPR.rnd = pr$rand$auc.integral,
					ROC = roc$auc,
					ROC.rnd = roc$rand$auc,
					FPR = sum(df$p.value[isZero] < 0.05) / sum(isZero))
			}
			res
		})
		do.call(rbind,res1)
	})
	df_aupr = do.call(rbind, df_aupr)

	df_return = cbind(df_aupr, n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, tau)

	df_return
}
```


# Simulations
```{r run_simulations, cache=TRUE}
run_three_formulas = function(n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, methods, tau, use.sccoda, mc.cores, short=TRUE){

	RNGkind("L'Ecuyer-CMRG")

	set.seed(1)
	formula = ~ trait
	res_none = run_simulation( n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, formula, methods, tau, use.sccoda, mc.cores)
	res_none$Status = "none"

	if( short ) return(res_none)

	set.seed(1)
	formula = ~ trait + Batch
	res_fixed = run_simulation( n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, formula, methods, tau, use.sccoda, mc.cores)
	res_fixed$Status = "fixed"
	  
	df_res = rbind(res_none, res_fixed)

	df_res$Status = factor(df_res$Status, c("none", "fixed", "random"))
	# return(df_res)
	   
	set.seed(1)
	formula = ~ trait + (1|Batch)
	res_random = run_simulation( n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, formula, methods, tau, FALSE, mc.cores)
	res_random$Status = "random"

	df_res = rbind(df_res, res_random)

	df_res$method = factor(df_res$method)
	df_res$Status = factor(df_res$Status, c("none", "fixed", "random"))

	df_res
} 
```

```{r model1}
mc.cores = ifelse(Sys.info()["sysname"] == "Darwin", 10, 44)
n_samples = 200
n_clusters = 10
mean_counts = 2000
n_sims = 10000 
beta.true = .2
n_batches = floor(n_samples/2)
var.batch = 0
tau = 10 
methods = c( "nb", "binomial", "betabinomial", "lm", "lmlog", "poisson")    
          
df_res = run_three_formulas( n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, methods, tau=tau, use.sccoda=FALSE, mc.cores=mc.cores, short=TRUE)  
```


# Fixed effects model
Parameters:

 - n_samples: `r n_samples`
 - n_clusters: `r n_clusters`
 - mean_counts: `r mean_counts`
 - beta.true: `r beta.true`
 - var.batch: `r var.batch`
 - tau: `r tau`


```{r plot.sims, cache=FALSE, fig.height=4, fig.width=11}
cols = brewer.pal(9, "Set1")
cols[cols=="#FFFF33"] = "#FDDB27"
names(cols) = c("crumblr", "CLR", "Linear model", "Linear model - log", "Binomial", "Beta-binomial", "NB", "Poisson", "scCoda")

CT = "CT_1"
# AUPR
aupr.rand = unique(df_res$AUPR.rnd[!is.na(df_res$AUPR.rnd)])
fig1 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
	ggplot(aes(method, AUPR, fill=method)) + 
	geom_bar(stat="identity", position = "dodge") + 
	theme_classic() + 
	coord_flip() + 
	scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) + 
	theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") + 
	scale_alpha_discrete( range = c(.3, 1)) + 
	ggtitle("AUPR") + 
	geom_hline(yintercept=aupr.rand, linetype="dashed", color="black", size=1) +
	scale_fill_manual(values=cols)

# FPR
fig2 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
		ggplot( aes(method, FPR, fill=method)) + 
		geom_bar(stat="identity", position = "dodge") + 
		theme_classic() + 
		coord_flip() + 
		scale_y_continuous(expand=c(0, 0), limits=c(0, NA)) + 
		theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") + 
		scale_alpha_discrete( range = c(.3, 1)) + ggtitle("False positive rate") + 
		geom_hline(yintercept=0.05, linetype="dashed", color="black", size=1) + 
		ylab("False positive rate") +
	scale_fill_manual(values=cols)

plot_grid(fig1, fig2, nrow=1, align="hv", axis="tblr")
```

# With scCoda
```{r scCoda, eval=FALSE}
n_samples = 200
n_clusters = 10
mean_counts = 2000
n_sims = 100 
beta.true = .2
n_batches = floor(n_samples/2)
var.batch = 0
tau = 10 
methods = c( "nb", "binomial", "betabinomial", "lm", "lmlog", "poisson")    
          
df_res = run_three_formulas( n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, methods, tau=tau, use.sccoda=TRUE, mc.cores=1, short=TRUE)  
```

Parameters:

 - n_samples: `r n_samples`
 - n_clusters: `r n_clusters`
 - mean_counts: `r mean_counts`
 - beta.true: `r beta.true`
 - var.batch: `r var.batch`
 - tau: `r tau`


```{r plot.sims.scoda, cache=FALSE, fig.height=4, fig.width=11, eval=FALSE}
cols = brewer.pal(9, "Set1")
cols[cols=="#FFFF33"] = "#FDDB27"
names(cols) = c("crumblr", "CLR", "Linear model", "Linear model - log", "Binomial", "Beta-binomial", "NB", "Poisson", "scCoda")

CT = "CT_1"
# AUPR
aupr.rand = unique(df_res$AUPR.rnd[!is.na(df_res$AUPR.rnd)])
fig1 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
	ggplot(aes(method, AUPR, fill=method)) + 
	geom_bar(stat="identity", position = "dodge") + 
	theme_classic() + 
	coord_flip() + 
	scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) + 
	theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") + 
	scale_alpha_discrete( range = c(.3, 1)) + 
	ggtitle("AUPR") + 
	geom_hline(yintercept=aupr.rand, linetype="dashed", color="black", size=1) +
	scale_fill_manual(values=cols)

# FPR
fig2 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
		ggplot( aes(method, FPR, fill=method)) + 
		geom_bar(stat="identity", position = "dodge") + 
		theme_classic() + 
		coord_flip() + 
		scale_y_continuous(expand=c(0, 0), limits=c(0, NA)) + 
		theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") + 
		scale_alpha_discrete( range = c(.3, 1)) + ggtitle("False positive rate") + 
		geom_hline(yintercept=0.05, linetype="dashed", color="black", size=1) + 
		ylab("False positive rate") +
	scale_fill_manual(values=cols)

plot_grid(fig1, fig2, nrow=1, align="hv", axis="tblr")
```



# Mixed model
```{r run.mixed}
n_batches = floor(n_samples/2)
var.batch = .1
  
df_res = run_three_formulas( n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, methods, tau=tau, use.sccoda=FALSE, mc.cores=mc.cores, short=TRUE)  
```

Parameters:

 - n_samples: `r n_samples`
 - n_clusters: `r n_clusters`
 - mean_counts: `r mean_counts`
 - beta.true: `r beta.true`
 - var.batch: `r var.batch`
 - tau: `r tau`

```{r plot.sims.mixed, cache=FALSE, fig.height=4, fig.width=11}
CT = "CT_1"
# AUPR
aupr.rand = unique(df_res$AUPR.rnd[!is.na(df_res$AUPR.rnd)])
fig1 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
		ggplot(aes(method, AUPR, fill=method, alpha=Status)) + 
		geom_bar(stat="identity", position = "dodge") + 
		theme_classic() + 
		coord_flip() + 
		scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") + 
		scale_alpha_discrete( range = c(1, 1)) + 
		ggtitle("AUPR") + 
		geom_hline(yintercept=aupr.rand, linetype="dashed", color="black", size=1) + 
		scale_fill_manual(values=cols)

# FPR
fig2 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
		ggplot(aes(method, FPR, fill=method, alpha=Status)) + 
		geom_bar(stat="identity", position = "dodge") + 
		theme_classic() + 
		coord_flip() + 
		scale_y_continuous(expand=c(0, 0), limits=c(0, NA)) + 
		theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") + 
		scale_alpha_discrete( range = c(1, 1)) + 
		ggtitle("False positive rate") + 
		geom_hline(yintercept=0.05, linetype="dashed", color="black", size=1) + ylab("False positive rate") + 
		scale_fill_manual(values=cols)

plot_grid(fig1, fig2, nrow=1, align="hv", axis="tblr")
```




# Increasing number of cells
```{r increase.cells}
n_batches = floor(n_samples/2)
var.batch = 0 
       
df_res = lapply( c(100, 500,1000, 2000, 4000), function(mean_counts){
	run_three_formulas( n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, methods, tau=tau, use.sccoda=FALSE, mc.cores=mc.cores) 
})
df_res = do.call(rbind, df_res)
```

Parameters:

 - n_samples: `r n_samples`
 - n_clusters: `r n_clusters`
 - beta.true: `r beta.true`
 - var.batch: `r var.batch`
 - tau: `r tau`

```{r plot.increasing, cache=FALSE, fig.width=11, fig.height=4}
CT = "CT_1"

# AUPR
aupr.rand = unique(df_res$AUPR.rnd[!is.na(df_res$AUPR.rnd)])
fig1 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
	ggplot(aes(factor(mean_counts), AUPR, fill=method)) +
	geom_bar(stat="identity", position = "dodge") +
	theme_classic() +
	scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) +
	 theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") +
	 geom_hline(yintercept=aupr.rand, linetype="dashed", color="black", size=1) +
	 xlab("Mean counts per sample") + 
	 scale_fill_manual(values=cols)

# FPR
fig2 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
	ggplot(aes(factor(mean_counts), FPR, fill=method)) +
	geom_bar(stat="identity", position = "dodge") +
	theme_classic() +
	scale_y_continuous(expand=c(0, 0), limits=c(0, NA)) +
	 theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
	 geom_hline(yintercept=0.05, linetype="dashed", color="black", size=1) +
	 xlab("Mean counts per sample") + 
	 scale_fill_manual(values=cols)

plot_grid(fig1, fig2, nrow=1, align="hv", axis="tblr", labels=LETTERS[1:2])
```



# Increasing overdispersion
```{r increase.tau}
df_res = lapply(c(1, 5, 10, 20), function(tau){
	run_three_formulas( n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, methods, tau=tau, use.sccoda=FALSE, mc.cores=mc.cores) 
}) 
df_res = do.call(rbind, df_res)
```


Parameters:

 - n_samples: `r n_samples`
 - n_clusters: `r n_clusters`
 - mean_counts: `r mean_counts`
 - beta.true: `r beta.true`
 - var.batch: `r var.batch`

```{r plot.increasing.tau, cache=FALSE, fig.width=11, fig.height=4}
CT = "CT_1"
# AUPR
aupr.rand = unique(df_res$AUPR.rnd[!is.na(df_res$AUPR.rnd)])
fig1 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
	ggplot(aes(factor(tau), AUPR, fill=method)) +
	geom_bar(stat="identity", position = "dodge") +
	theme_classic() +
	scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) +
	 theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") +
	 geom_hline(yintercept=aupr.rand, linetype="dashed", color="black", size=1) +
	 xlab(bquote(Overdispersion~tau)) + 
	 scale_fill_manual(values=cols)

# FPR
fig2 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
	ggplot(aes(factor(tau), FPR, fill=method)) +
	geom_bar(stat="identity", position = "dodge") +
	theme_classic() +
	scale_y_continuous(expand=c(0, 0), limits=c(0, NA)) +
	 theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
	 geom_hline(yintercept=0.05, linetype="dashed", color="black", size=1) +
	 xlab(bquote(Overdispersion~tau)) + 
	 scale_fill_manual(values=cols)

plot_grid(fig1, fig2, nrow=1, align="hv", axis="tblr", labels=LETTERS[1:2])
```



# Number of components
```{r increase.clusters} 
df_res = lapply(c(7, 10, 15, 20), function(n_clusters){
	run_three_formulas( n_samples, n_clusters, mean_counts=mean_counts, n_sims, beta.true, n_batches, var.batch, methods, tau, use.sccoda=FALSE, mc.cores=mc.cores) 
}) 
df_res = do.call(rbind, df_res)
```


Parameters:

 - n_samples: `r n_samples`
 - mean_counts: `r mean_counts`
 - beta.true: `r beta.true`
 - var.batch: `r var.batch`
 - tau: `r tau`

```{r plot.increase.clusters, cache=FALSE, fig.width=11, fig.height=4}
CT = "CT_1"

# AUPR
aupr.rand = unique(df_res$AUPR.rnd[!is.na(df_res$AUPR.rnd)])
fig1 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
	ggplot(aes(factor(n_clusters), AUPR, fill=method)) +
	geom_bar(stat="identity", position = "dodge") +
	theme_classic() +
	scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) +
	 theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") +
	 geom_hline(yintercept=aupr.rand, linetype="dashed", color="black", size=1) +
	 xlab("Number of components") + 
	 scale_fill_manual(values=cols)

# FPR
fig2 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
	ggplot(aes(factor(n_clusters), FPR, fill=method)) +
	geom_bar(stat="identity", position = "dodge") +
	theme_classic() +
	scale_y_continuous(expand=c(0, 0), limits=c(0, NA)) +
	 theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
	 geom_hline(yintercept=0.05, linetype="dashed", color="black", size=1) +
	 xlab("Number of components") + 
	 scale_fill_manual(values=cols)

plot_grid(fig1, fig2, nrow=1, align="hv", axis="tblr", labels=LETTERS[1:2])
```



# Sample size
```{r increase.samples} 
df_res = lapply(c(20, 50, 100, 200), function(n_samples){
	run_three_formulas( n_samples, n_clusters, mean_counts, n_sims, beta.true, n_batches, var.batch, methods, tau, use.sccoda=FALSE, mc.cores=mc.cores) 
}) 
df_res = do.call(rbind, df_res)
```


Parameters:

 - n_clusters: `r n_clusters`
 - mean_counts: `r mean_counts`
 - beta.true: `r beta.true`
 - var.batch: `r var.batch`
 - tau: `r tau`


```{r plot.increase.samples, cache=FALSE, fig.width=11, fig.height=4}
CT = "CT_1"
# AUPR
aupr.rand = unique(df_res$AUPR.rnd[!is.na(df_res$AUPR.rnd)])
fig1 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
	ggplot(aes(factor(n_samples), AUPR, fill=method)) +
	geom_bar(stat="identity", position = "dodge") +
	theme_classic() +
	scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) +
	 theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") +
	 geom_hline(yintercept=aupr.rand, linetype="dashed", color="black", size=1) +
	 xlab("Number of samples") + 
	 scale_fill_manual(values=cols)

# FPR
fig2 = df_res %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
	ggplot(aes(factor(n_samples), FPR, fill=method)) +
	geom_bar(stat="identity", position = "dodge") +
	theme_classic() +
	scale_y_continuous(expand=c(0, 0), limits=c(0, NA)) +
	 theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
	 geom_hline(yintercept=0.05, linetype="dashed", color="black", size=1) + 
	 xlab("Number of samples") + 
	 scale_fill_manual(values=cols)

plot_grid(fig1, fig2, nrow=1, align="hv", axis="tblr", labels=LETTERS[1:2])
```




# Increasing batch effect
```{r sim2}  
n_batches = 50 
 
df_res2 = lapply(c(0, .05, .1, .2), function(var.frac){  

	var.batch = beta.true^2* var.frac / (1 - var.frac)

	run_three_formulas( n_samples, n_clusters, mean_counts, n_sims = n_sims, beta.true, n_batches, var.batch, methods = methods, tau=tau, use.sccoda=FALSE, mc.cores=1, short=FALSE) 
}) 
df_res2 = do.call(rbind, df_res2)
```


Parameters:

 - n_samples: `r n_samples`
 - n_clusters: `r n_clusters`
 - mean_counts: `r mean_counts`
 - beta.true: `r beta.true`
 - n_batches: `r n_batches`
 - tau: `r tau`


```{r sim2.plots, fig.height=4, fig.width=11, cache=FALSE, eval=FALSE}
aupr.rand = unique(df_res2$AUPR.rnd[!is.na(df_res$AUPR.rnd)])
df_res2 %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
		ggplot(aes(method, AUPR, fill=method, alpha=Status)) + geom_bar(stat="identity", position = "dodge") + theme_classic() + coord_flip() + facet_wrap(~mean_counts + var.batch, ncol=4) + scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="right")  + geom_hline(yintercept=aupr.rand, linetype="dashed", color="black", size=1) + scale_alpha_discrete( range = c(.3, 1)) + scale_fill_manual(values=cols)

ymax = max(max(df_res2$FPR), .1)
df_res2 %>%
		as_tibble %>%
		filter(assay == CT) %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
		ggplot(aes(method, FPR, fill=method, alpha=Status)) + geom_bar(stat="identity", position = "dodge") + theme_classic() + coord_flip() + facet_wrap(~mean_counts + var.batch, ncol=4) + scale_y_continuous(expand=c(0, 0), limits=c(0, ymax)) + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="right")  + geom_hline(yintercept=0.05, linetype="dashed", color="black", size=1) + scale_alpha_discrete( range = c(.3, 1)) + scale_fill_manual(values=cols)
```



```{r plot.batch.effect, cache=FALSE, fig.width=11, fig.height=8}
CT = "CT_1"
# AUPR
aupr.rand = unique(df_res2$AUPR.rnd[!is.na(df_res2$AUPR.rnd)])

fig1 = df_res2 %>%
	filter(assay==CT) %>%
	# filter(Status != "none") %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
		mutate(var.frac = format(100 * var.batch / (var.batch + beta.true^2), digits=2)) %>%		 
 	ggplot(aes(var.frac, AUPR, fill=method)) +
	geom_bar(stat="identity", position = "dodge") +
	theme_classic() +
	scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) +
	 theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") +
	 geom_hline(yintercept=aupr.rand, linetype="dashed", color="black", size=1) +
	 xlab("Batch effect") + 
	 scale_fill_manual(values=cols) +
	 facet_wrap(~Status)

# FPR
fig2 = df_res2 %>%
	filter(assay==CT) %>%
	# filter(Status != "none") %>%
		mutate(method = recode(method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>%
		mutate(var.frac = format(100 * var.batch / (var.batch + beta.true^2), digits=2)) %>%		 
	ggplot(aes(var.frac, FPR, fill=method)) +
	geom_bar(stat="identity", position = "dodge") +
	theme_classic() +
	scale_y_continuous(expand=c(0, 0), limits=c(0, NA)) +
	 theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
	 geom_hline(yintercept=0.05, linetype="dashed", color="black", size=1) + 
	 xlab("Batch effect") + 
	 scale_fill_manual(values=cols)+
	 facet_wrap(~Status)

plot_grid(fig1, fig2, ncol=1, align="hv", axis="tblr", labels=LETTERS[1:2])
```




# Method times
```{r times}
# n_samples = 500
# n_clusters = 20
RhpcBLASctl::omp_set_num_threads(1)
RhpcBLASctl::blas_set_num_threads(1)
 
# each batch is observed at last twice
info = data.frame(trait = rnorm(n_samples), Batch = factor(rep(sample.int(n_batches, n_samples/2, replace=TRUE), 2)))
rownames(info) = paste0("sample_", 1:nrow(info))

value = beta.true
beta = data.frame(mu = rep(0, n_clusters), b = c(0, value, rep(0, n_clusters-2)))
a = n_clusters 
b = nlevels(info$Batch)-1
v = rgamma(1, var.batch, 1)
beta_batch = matrix(rnorm(a*b, sd=sqrt(v)), a, b)
beta = cbind(beta, beta_batch)

eta = model.matrix(~., info) %*% t(beta)

probs = exp(eta)

totalCounts = rnegbin(nrow(probs), mu=mean_counts, theta = 3)

if( tau == 1){
	# large a0 corresponds to multinomial
	alpha = probs * 1e9
}else{
	# convert tau overdispersion value to alpha from p
	rhoSq = (tau - 1) / (mean(totalCounts)-1)
	a0 = (1-rhoSq) / rhoSq
	alpha = probs*a0
}

# Multinomial-Dirichlet
counts = lapply( 1:length(totalCounts), function(i){
	Dirichlet.multinomial(totalCounts[i], alpha[i,]) 
})
counts = do.call(rbind, counts)
rownames(counts) = paste0("sample_", 1:nrow(counts))
colnames(counts) = paste0("CT_", 1:ncol(counts))

cellTypeFreq = rowMeans(apply(counts, 1, function(x) x/sum(x)))
df_cellTypeFreq = data.frame(cellTypeFreq)

formula = ~ trait + Batch

df_time = data.frame()

# CLR
res = system.time(replicate(100, {
Y = t(as.data.frame(compositions::clr(counts)))
fit_clr = suppressMessages(dream( Y, formula, info, BPPARAM=SerialParam(), useWeights=FALSE))
fit_clr = eBayes(fit_clr)
tab_clr = topTable(fit_clr, coef="trait", sort.by="none")
}))
df_time = rbind(df_time, data.frame(Method = "CLR", Time = as.numeric(res[3]) / 100))


# crumblr
res = system.time(replicate(100,{
cobj = crumblr(counts)
fit = suppressMessages(dream(cobj, formula, info, BPPARAM=SerialParam()))
fit = eBayes(fit)
tab.crumblr = topTable(fit, coef="trait", sort.by="none")
}))
df_time = rbind(df_time, data.frame(Method = "crumblr", Time = as.numeric(res[3]) / 100))

# scCoda
res = system.time({
		a = sccoda_wrapper(counts, info, formula, python_path = "/sc/arion/work/hoffmg01/condaEnv/sccoda/bin/python")
})
df_time = rbind(df_time, data.frame(Method = "scCoda", Time = as.numeric(res[3])))

df_time_multiple = lapply( methods, function(method){

		res = system.time(replicate(5, {
			# keep only samples with at elast 10 reads
			i = rowSums(counts) > 10
			res = testComposition( counts[i,], formula, info[i,], coef="trait", eval="test", method = method)
		}))
		data.frame(Method = method, Time = as.numeric(res[3])/5)
})
df_time = rbind(df_time, do.call(rbind, df_time_multiple))
```

```{r plot.time, fig.width=4, fig.height=4, cache=FALSE}
ymax = log10(max(df_time$Time)) + 2

df_time %>%
	mutate(Method = recode(Method, poisson = "Poisson", nb = "NB", lm="Linear model", lmlog = "Linear model - log", binomial="Binomial", betabinomial="Beta-binomial")) %>% 
 ggplot(aes(Method, log10(Time)+2, fill=Method)) +
	geom_bar(stat="identity") +
	theme_bw() +
	theme(aspect.ratio=1, legend.position="none", panel.grid.minor = element_blank(), panel.grid.major.y = element_blank()) +
	scale_y_continuous(bquote(Run~time~(seconds~log[10]~scale)), limits=c(0, ymax*1.05), expand=c(0,0), breaks=(0:5), labels=10^(0:5  -2))  + 
	coord_flip() + 
	scale_fill_manual(values=cols)
```










# Session Info
<details>
```{r session_info}
sessionInfo()
```
<\details>




