---
title: "Analysis of Alzheimer's Disease in [PsychAD](https://adknowledgeportal.synapse.org/Explore/Projects/DetailsPage?Grant%20Number=R01AG067025)"
subtitle: 'Public Release 0'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
output: 
  html_document:
    toc: true
    smart: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  package.startup.message = FALSE,
  cache = TRUE,
  cache.lazy = FALSE)
```

# Load libraries
```{r load.packages, cache=FALSE}
suppressPackageStartupMessages({
library(SingleCellExperiment)
library(zellkonverter)
library(dreamlet)
library(crumblr)
library(aplot)
library(tidyverse)
library(kableExtra)
library(ggplot2)
library(ggtree)
library(ggcorrplot)
library(cowplot)
library(scattermore)
})
```


```{r load.data}
folder = "/sc/arion/projects/psychAD/NPS-AD/public_release_0/" 
file = paste0(folder, "PsychAD_r0_Dec_28_2022.h5ad")
sce = readH5AD(file, use_hdf5=TRUE, verbose=TRUE)
assayNames(sce)[1] = "counts"
sce$Dx = factor(sce$Dx_AD, c('Control','AD'))

sce$pool = gsub("^(.*)-\\S+$", "\\1", sce$poolID)

pb.subclass <- aggregateToPseudoBulk(sce,
    assay = "counts",     
    cluster_id = "subclass",  
    sample_id = "Channel", 
    BPPARAM = SnowParam(6, progressbar=TRUE))

pb.subtype <- aggregateToPseudoBulk(sce,
    assay = "counts",     
    cluster_id = "subtype",  
    sample_id = "Channel", 
    BPPARAM = SnowParam(6, progressbar=TRUE))

pb.subclass$pH[is.na(pb.subclass$pH)] = mean(pb.subclass$pH, na.rm=TRUE)
pb.subtype$pH[is.na(pb.subtype$pH)] = mean(pb.subtype$pH, na.rm=TRUE)
```

Public freeze 0 includes `r length(table(sce$Channel))` samples, `r length(table(sce$round_num))` rounds, `r length(table(sce$poolID))` 10X batches, `r length(table(sce$SubID))` donors, and `r format(ncol(sce), big.mark=',')` cells passing QC.


```{r umap.subclass}
# extract UMAP coordinates and annotations
df = cbind(reducedDim(sce, "X_umap"), 
    colData(sce)[,c("subclass", "subtype")]) %>% 
    as.data.frame

ggplot(df, aes(V1, V2, color=subclass)) + geom_scattermore() + theme_classic() + theme(aspect.ratio=1) + guides(colour = guide_legend(override.aes = list(size = 1.5))) + xlab("UMAP1") + ylab("UMAP2") 
```


# Subclass
```{r cca}
form = ~ SubID + poolID + pool + Sex + scale(Age) + Dx + pH + PMI
C = canCorPairs(form, colData(pb.subclass) )
ggcorrplot(C, hc.order = TRUE)
```

## Plots of cell fractions
```{r fractions, fig.width=9}
fracs = cellCounts(pb.subclass) / rowSums(cellCounts(pb.subclass))

i = pb.subclass$Dx == 'Control'
fig1 = plotPercentBars(fracs[i,], col=ggColorHue(ncol(fracs))) + 
  ylab("Cell fractions") + 
  theme(legend.position = "none", axis.text.y=element_blank())

i = pb.subclass$Dx == "AD"
fig2 = plotPercentBars(fracs[i,], col=ggColorHue(ncol(fracs))) + 
  ylab("Cell fractions") + 
  theme(legend.position = "none", axis.text.y=element_blank())

plot_grid(fig1, fig2, rel_widths=c(.75,1))
```

```{r plot.fracs, fig.width=8, fig.height=9}
df = data.frame(fracs, diagnosis = pb.subclass$Dx, check.names=FALSE)
df = reshape2::melt(df, id.vars="diagnosis")
 
ggplot(df, aes(diagnosis, value, fill=variable)) + 
  geom_violin() +
  geom_boxplot(fill="grey50", width=.1) + 
  facet_wrap(~ variable) +
  theme_classic() +
  theme(aspect.ratio=1, legend.position="none") +
  ylab(bquote(Fraction~(log[10]))) +
  scale_y_log10()
```

```{r fraction.table}
df %>%
  group_by(variable, diagnosis) %>%
  # mutate(diagnosis = recode(diagnosis, '0' = "Control", '1' = "Alzheimer's"))  %>% 
  summarize(mean = mean(value)) %>%
  pivot_wider(names_from=c('diagnosis'), values_from=c("mean")) %>%
  kbl %>%
  kable_styling(full_width = F)
```



## crumblr
### Variance partitioning analysis
```{r crumblr}
cobj = crumblr(cellCounts(pb.subclass))

form = ~ (1|SubID) + (1|pool) + (1|Sex) + scale(Age) + (1|Dx) + pH + PMI

vp.c = fitExtractVarPartModel(cobj, form, colData(pb.subclass))

plotVarPart(sortCols(vp.c), label.angle=60, ncol=4) + theme(aspect.ratio=1)
```

```{r plotPercentBars}
fig.vp = plotPercentBars( sortCols(vp.c) )

# analysis with dream()
form = ~ (1|SubID) + (1|pool) + (1|Sex) + scale(Age) + Dx + pH + PMI
fit = dream( cobj, form, colData(pb.subclass))
fit = eBayes(fit)
topTable(fit, coef="DxAD")
```

# Age in controls
```{r age}
# cn = colnames(cellCounts(pb.subclass))

# i = which(cn %in% c("Olig", "OPC", "Astro"))
# i = grep("IN|EN", cn)

pb.controls = pb.subclass[,pb.subclass$Dx == "Control"]

cobj = crumblr(cellCounts(pb.controls))
form = ~ (1|SubID) + (1|pool) + Sex + Age + pH + PMI

fit = dream( cobj, form, colData(pb.controls))
fit = eBayes(fit)

coef = 'Age'
topTable(fit, coef=coef)
```

### Multivariate test along hierarchy
```{r sd, fig.width=10, fig.height=5}
hc = buildClusterTreeFromPB(pb.controls)

res = treeTest( fit, cobj, hc, coef=coef)

fig1 = plotTreeTest(res) + xlim(0, 15) + theme(legend.position="none")

tab = topTable(fit, coef=coef, number=Inf)

tab$celltype = factor(rownames(tab), rev(get_taxa_name(fig1)))
tab$se = with(tab, logFC/t)

fig2 = ggplot(tab, aes(celltype, logFC)) + 
  geom_hline(yintercept=0, linetype="dashed", color="grey", size=1) +
  geom_errorbar(aes(ymin = logFC - 1.96*se, ymax = logFC + 1.96*se), width=0) +
  geom_point(color="dodgerblue") +
  theme_classic() +
  coord_flip() +
  xlab('') + 
  ylab("Effect size") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

# combine plots
fig2 %>% insert_left(fig1) %>% insert_right(fig.vp)
```



### Correlation between coefficient estimates
```{r ggcorrplot}
# get covariance matrix
V = vcov(fit, cobj, coef=coef)

# convert to correlation matrix
V = cov2cor(V)

# shorten names
rownames(V) = gsub(paste0(":", coef), '', rownames(V))
colnames(V) = gsub(paste0(":", coef), '', colnames(V))

# sort to match tree
i = match(rev(get_taxa_name(fig1)), rownames(V))

ggcorrplot(V[i,i], outline.color=NA, colors=c("blue3", "white", "red3"))
```

How to compute residuals.  Ask Roman

```{r plots}

df = data.frame(IN_SST = cobj$E["IN_SST",],
                Age = pb.controls$Age,
                se = 1/sqrt(cobj$weights["IN_SST",]))

# png("~/www/test.png")
ggplot(df, aes(Age, IN_SST)) +
  geom_errorbar(aes(ymin = IN_SST - 1.96*se, ymax = IN_SST + 1.96*se), width=0, color="grey50") +
  geom_point() +
  theme_classic() +
  theme(aspect.ratio=1) +
  geom_smooth(method="lm")
# dev.off()


fracs = cellCounts(pb.controls) / rowSums(cellCounts(pb.controls))


df = data.frame(IN_SST = fracs[,"IN_SST"], 
                Age = pb.controls$Age)

# png("~/www/test.png")
ggplot(df, aes(Age, IN_SST)) +
  geom_point() +
  theme_classic() +
  theme(aspect.ratio=1) +
  geom_smooth(method="lm")
# dev.off()





```



```{r exit, cache=FALSE}
knitr::knit_exit(1)
```




# Subtype

```{r umap.subtype}
# extract UMAP coordinates and annotations
df = cbind(reducedDim(sce, "X_umap"), 
    colData(sce)[,c("subclass", "subtype")]) %>% 
    as.data.frame

ggplot(df, aes(V1, V2, color=subtype)) + geom_scattermore() + theme_classic() + theme(aspect.ratio=1) + guides(colour = guide_legend(override.aes = list(size = 1.5))) + xlab("UMAP1") + ylab("UMAP2") 
```


```{r cca.subtype}
form = ~ SubID + poolID + pool + Sex + scale(Age) + Dx + pH + PMI
C = canCorPairs(form, colData(pb.subtype) )
ggcorrplot(C, hc.order = TRUE)
```

## Plots of cell fractions
```{r fractions.subtype, fig.width=9}
fracs = cellCounts(pb.subtype) / rowSums(cellCounts(pb.subtype))

i = pb.subtype$Dx == "Control"
fig1 = plotPercentBars(fracs[i,], col=ggColorHue(ncol(fracs))) + 
  ylab("Cell fractions") + 
  theme(legend.position = "none", axis.text.y=element_blank())

i = pb.subtype$Dx == "AD"
fig2 = plotPercentBars(fracs[i,], col=ggColorHue(ncol(fracs))) + 
  ylab("Cell fractions") + 
  theme(legend.position = "none", axis.text.y=element_blank())

plot_grid(fig1, fig2, rel_widths=c(.75,1))
```




```{r plot.fracs.subtype, fig.width=8, fig.height=9}
df = data.frame(fracs, diagnosis = pb.subtype$Dx, check.names=FALSE)
df = reshape2::melt(df, id.vars="diagnosis")
 
ggplot(df, aes(diagnosis, value, fill=variable)) + 
  geom_violin() +
  geom_boxplot(fill="grey50", width=.1) + 
  facet_wrap(~ variable) +
  theme_classic() +
  theme(aspect.ratio=1, legend.position="none") +
  ylab(bquote(Fraction~(log[10]))) +
  scale_y_log10()
```


```{r fraction.table.subtype}
df %>%
  group_by(variable, diagnosis) %>%
  # mutate(diagnosis = recode(diagnosis, '0' = "Control", '1' = "Alzheimer's"))  %>% 
  summarize(mean = mean(value)) %>%
  pivot_wider(names_from=c('diagnosis'), values_from=c("mean")) %>%
  kbl %>%
  kable_styling(full_width = F)
```






## crumblr
### Variance partitioning analysis
```{r crumblr.subtype}
cobj = crumblr(cellCounts(pb.subtype))

form = ~ (1|SubID) + (1|poolID) + (1|Sex) + scale(Age) + (1|Dx) + pH + PMI

vp.c = fitExtractVarPartModel(cobj, form, colData(pb.subtype))

plotVarPart(sortCols(vp.c), label.angle=60, ncol=4)  + theme(aspect.ratio=1)
```

```{r plotPercentBars.subtype}
fig.vp = plotPercentBars( sortCols(vp.c) )

# analysis with dream()
form = ~ (1|SubID) + (1|poolID) + (1|Sex) + scale(Age) + Dx + pH + PMI
fit = dream( cobj, form, colData(pb.subtype))
fit = eBayes(fit)
```



### Multivariate test along hierarchy
```{r sd.subtype, fig.width=12, fig.height=7}
hc = buildClusterTreeFromPB(pb.subtype)

res = treeTest( fit, cobj, hc, coef="DxAD")

fig1 = plotTreeTest(res) + xlim(0, 15) + theme(legend.position="none")

tab = topTable(fit, coef="DxAD", number=Inf)

tab$celltype = factor(rownames(tab), rev(get_taxa_name(fig1)))
tab$se = with(tab, logFC/t)

fig2 = ggplot(tab, aes(celltype, logFC)) + 
  geom_hline(yintercept=0, linetype="dashed", color="grey", size=1) +
  geom_errorbar(aes(ymin = logFC - 1.96*se, ymax = logFC + 1.96*se), width=0) +
  geom_point(color="dodgerblue") +
  theme_classic() +
  coord_flip() +
  xlab('') + 
  ylab("Effect size") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

# combine plots
fig2 %>% insert_left(fig1) %>% insert_right(fig.vp)
```


### Correlation between coefficient estimates
```{r ggcorrplot.subtype, fig.width=9, fig.height=9}
# get covariance matrix
V = vcov(fit, cobj, coef="DxAD")

# convert to correlation matrix
V = cov2cor(V)

# shorten names
rownames(V) = gsub(":DxAD", '', rownames(V))
colnames(V) = gsub(":DxAD", '', colnames(V))

# sort to match tree
i = match(rev(get_taxa_name(fig1)), rownames(V))

ggcorrplot(V[i,i], outline.color=NA, colors=c("blue3", "white", "red3"))
```


# Session Info
<details>
```{r sessioninfo, cache=FALSE}
sessionInfo()
```
</details>
