---
title: "Analysis of Alzheimer's Disease in [PsychAD](https://adknowledgeportal.synapse.org/Explore/Projects/DetailsPage?Grant%20Number=R01AG067025)"
subtitle: 'Public Release 0'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
output: 
  html_document:
    toc: true
    smart: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  package.startup.message = FALSE,
  cache = TRUE,
  cache.lazy = FALSE)
```

# Load libraries
```{r load.packages, cache=FALSE}
suppressPackageStartupMessages({
library(SingleCellExperiment)
library(zellkonverter)
library(dreamlet)
library(crumblr)
library(aplot)
library(tidyverse)
library(kableExtra)
library(ggplot2)
library(scattermore)
})
```


```{r load.data}
 # Public freeze 0
h5ad_file = "/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/h5ad/221202_NPS-AD_public0_pass3_anno_clean.h5ad"
sce = readH5AD(file, use_hdf5=TRUE, raw=TRUE, verbose=TRUE, uns=FALSE)

# drop cells with annotation "Unknown"
sce = sce[,sce$celltype != "Unknown"]
colData(sce) = droplevels(colData(sce)) 

# AD status as factor
sce$AD = factor(sce$AD, 0:1)
```

Public freeze 0 includes `r length(table(sce$Channel))` samples, `r length(table(sce$round_num))` rounds, `r length(table(sce$batch))` 10X batches, `r length(table(sce$SubID))` donors, and `r format(ncol(sce), big.mark=',')` cells passing QC.


```{r umap9, dev="png"}
# extract UMAP coordinates and annotations
df = cbind(reducedDim(sce, "X_umap"), 
    colData(sce)[,c("anno", "celltype", "class", "subtype")]) %>% 
    as.data.frame

ggplot(df, aes(V1, V2, color=subtype)) + geom_scattermore() + theme_classic() + theme(aspect.ratio=1) + guides(colour = guide_legend(override.aes = list(size = 1.5))) + xlab("UMAP1") + ylab("UMAP2") 
```


# Cell type composition
```{r test.crumblr, fig.width=8}
# Compute hierarchical clustering
hcl = buildClusterTree(sce, "X_pca_regressed", "subtype")

# Get cell counts
counts = computeCellCounts(sce, "subtype", "Channel")

idx = match( rownames(counts), sce$Channel)
info = colData(sce)[idx,]
rownames(info) = info$Channel

cobj = crumblr(counts)
```

## Variance partitioning analysis
```{r vp}
form = ~ (1|SubID) + (1|batch) + (1|Sex) + scale(Age) + (1|AD)

vp = fitExtractVarPartModel(cobj, form, info)

plotVarPart(sortCols(vp)) + theme(aspect.ratio=1)
```

## Differential testing
```{r diff.test}
form = ~ (1|SubID) + (1|batch) + (1|Sex) + scale(Age) + AD

fit = dream(cobj, form, info)
fit = eBayes(fit)

topTable(fit, coef='AD1', number=Inf) %>%   
  select(logFC, AveExpr, t, P.Value, adj.P.Val) %>% 
  kbl() %>%  
  kable_classic(full_width = FALSE)


## Differential testing using hierarchical clustering
```{r tree.test, fig.width=9}
res1 = treeTest( fit, cobj, hcl, coef="AD1", method="FE")
fig.tree = plotTreeTest(res1) + xlim(0, 13) + ggtitle("FE")

# plot of LogFC
tab = topTable(fit, "AD1", number=Inf, sort.by="none")

# sort based on tree
df = data.frame(ID = rownames(tab), tab)
df$se = with(df, logFC/ t)

fig.logFC = ggplot(df, aes(ID, logFC)) +
  geom_hline(yintercept=0, linetype="dashed", color="grey50") + 
  geom_errorbar(aes(ymin=logFC - 1.96*se, ymax=logFC + 1.96*se), width=0) +
  geom_point(color="dodgerblue") +
  coord_flip() +
  theme_classic() +
  xlab('') +
  theme(aspect.ratio=4, axis.text.y = element_blank())

fig.logFC %>% insert_left(fig.tree)
```

# Session Info
<details>
```{r sessioninfo, cache=FALSE}
sessionInfo()
```
</details>
